Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-04-16 Delaneyetal2024.tex
push to overleaf

\begin{keypoints}
\item Short water discharge fluctuations in subglacial channels are accommodated by water velocity, not channel size as in subaerial systems.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Significant subglacial sediment transport capacity can persist across a range of water discharges, impacting the  timing of peak events.
\end{keypoints}

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated  by flow depth and width changes, along with water velocity.
  However, in subglacial channels  water discharge changes mainly result in velocity variations, as the channel geometry evolves slowly (over days).
  Here, hydrographs from an Alpine glacier and the Greenland Ice Sheet are applied to subglacial and subaerial hydraulics models, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results also show that high subglacial sediment transport rates and water velocities can occur across wide-ranging water discharges.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
\end{abstract}

@@

Over long periods, such as millennia, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, should adequate sediment be present subglacially.

@@

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \cite{leopold1953}.
The change in water depth results in a proportional increase in water velocity.
As a result, the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels is controlled on the opposing processes of channel opening by frictional heating of water flow on the one hand versus creep closure by ice flow on the other \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days or longer, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, on short time scales, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated by increased or decreased water velocity $v$ \cite<Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; >{alley1997}.

@@

Sediment mobilization in subaerial and subglacial channels, therefore, responds differently to changing water discharge.
These differences are implicitly included in a wide range of available models that  quantify sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The different response of  sediment transport capacity to changing water discharge  may well impact sediment dynamics near glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming no water storage in the proglacial area.
The Alpine site (labeled \alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014, to October 10, 2014 \cite{felix2021,felix2022}.

The Leverett Glacier in Greenland (labeled \icesheet{}) serves as the ice sheet setting.
Water discharge was measured roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used in \citeA{tedstone2013} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.

@@

\section{Methods}
\label{sect:meth}
The models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of shear stress omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948}.

@@

Here, it is assumed that the water is transported through a subglacial channel \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972} beneath a glacier with channel length $l$, with a flatbed and a mean thickness of $h_{ice}$.
The channel  size grows from melt due to frictional heating from water flow and closes due to the creep  of the ice.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $t$ is time, $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (complete values given in Table \ref{table:vpm}), $g$ is the acceleration due to gravity, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change over $l$, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the mean ice overburden pressure expressed in meter water equivalent ($\rho_w$ is density of water; $\rho_i$ is density of ice), and $n$ is Glen's n \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_r\,\frac{v_{g}^{2}}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the channel length, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
%
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle, \citeA{hooke1990}). Note that $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
This completes the subglacial hydraulic model which is described by the state variables $S_g$ and $\Delta h$.

@@

The models above are applied to proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}).
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by Fieschergletscher's relatively thin ice thickness \cite<$h_{ice}$: $225$\,\unit{m}>{grab2021}, low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by Leverett glacier's  thick ice  \cite<$h_{ice}$: $700$\,\unit{m}; >{morlighem2017}, high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
%However, both scenarios share a discharge with a similar, relative seasonal amplitude.

In the first experiment, the models are  applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{6}$%, i.e. a segment of a quarter-circle,
(Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities \cite<$\sim\,1.6\,$\unit{m}\,\unit{s}$^{-1}$>{werder2010b,chandler2013} occur for both \alpine{} and \icesheet{}.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with the maximum observed water discharge for the first $4$ days of the study until there is no change in channel area. 
The subaerial channels of both scenarios have a bed slope of $\nabla z_c = 0.05$ (Table~\ref{table:vpm}).

The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
To quantify the link between sediment transport capacity and water discharge, model outputs are compared to water discharge from the glaciers using Spearman rank correlation, which varies with the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.

A second experiment aims to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels across a range of channel shapes, friction factors, and water discharge variability. 
To accomplish this, we examine the variability of one model output (i.e water velocity) by evaluating its standard deviation throughout the season.

The effect of channel shape and roughness is established by running the model with random parameter values of channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_f$; see Table~\ref{table:vpm} for value range).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}\,\unit{s}$^{-1}$ and $2$\,\unit{m}\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}\,\unit{s}$^{-1}$, following measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
Additionally, to be accepted, model runs can only experience a flotation fraction ($\frac{\Delta h}{2\,h_o}$) of above $1.2$ for less than $2.5$\,\% of the run.

The routine runs until $100$ different parameter combinations for each water discharge aggregation period are accepted with the conditions described above.
Here, the water discharge input at a given time is the average over the aggregation period before that time, ranging from $15$\,\unit{min} up to $15$\,\unit{days}.

\section{Results}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows denote where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the  peak melt event denoted by the shaded area in panels e--h, with an arbitrary y-axis.
    % Note that the x and y axes are different for \alpine{} and \icesheet{}.
  }
  \label{fig:model_outs}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig3.pdf}
  \caption{
    Relationship between water discharge and normalized velocity, shear stress, and width-integrated shear stress for \alpine{} (a-c) and \icesheet{} (d-f).
    Variables on the x-axis have been normalized to mean values.
    $R_g$ ($R_f$) shows the Spearman rank correlations for the subglacial (subaerial) outputs.
    Plots are shown with discharge and outputs at $15$\unit{min} intervals.
  }
  \label{fig:Qw_vari}
\end{figure}

Examining the relationship between water discharge and the model outputs, different relationships emerge between the subaerial and subglacial cases due to the hysteresis in channel size in the subglacial model (Equation~\ref{eq:dS_dt}).
Because there is no history dependence in  subaerial channels, each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari}).

Conversely, the history dependence on channel size in subglacial channels means different outputs occur across a range of discharges.
For instance, in subglacial channels in \alpine{}, very high seasonal water velocity values and shear stresses can occur from a low water discharge  ($\sim\,4$\,\unit{m}$^3$\,\unit{s}$^{-1}$) to the maximum water discharge at over ($17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$, Figure~\ref{fig:Qw_vari}\,a).
In \icesheet{}, water velocities close to the seasonal mean value can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
When considering the subglacial channel's evolving width, width-integrated shear stress  across the channel generally increases with water discharge, with improved rank correlation compared to water velocity or shear stress (\alpine{}, Figure~\ref{fig:Qw_vari} a--c).
Yet even the width-integrated shear stress  can vary substantially, with the highest values occurring at water discharge values ranging from roughly $11$ \,\unit{m}\,\unit{s}$^{-1}$ to over $17$ \,\unit{m}\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet{} scenario, where the hydrograph has less diurnal variability (Figure~\ref{fig:Qw_vari} c, f).

The different response of subaerial and subglacial model outputs to water discharge impacts the timing of peaks in sediment transport capacity.
In the subglacial channel, peaks in $w_g\tau_g$ generally occur when water discharge increases at the fastest rate, but before the maximum water discharge (Figure~\ref{fig:model_outs}\, d and h).
By contrast, in the subaerial channel,   $w_f\tau_f$  peaks when water discharge is highest.

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\linewidth]{Fig4.pdf}
    \caption{Standard deviation of model outputs for different time aggregations with different subglacial and subaerial channel shapes and friction factors.
      Shaded areas denote the range of results from the accepted $100$ parameter combinations (Section~\ref{sect:imp}).
      Solid lines denote  mean values.
      Markers show aggregation periods ($15$\,\unit{min}, $1$\,\unit{hr}, $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d}).
       }
    \label{fig:multi_run}
  \end{figure}

The second numerical experiment aims to evaluate the role of channel shape and friction factors on the findings above and to examine the effect of variability in water discharge.
Across the range of channel shapes and friction factors  examined (Section\,\ref{sect:imp}), variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one in both \alpine{} and \icesheet{}.
This suggests that there is greater intermittency and variability in sediment transport in subglacial systems compared to subaerial ones, especially in supply-limited systems where sediment could be exhausted or mobilized over diurnal discharge cycles.
Considering the width integrated shear stress, in \alpine{} the variability is comparable between the subaerial and subglacial systems, due to the highly variable water discharge, largely due to diurnal variations (Figure~\ref{fig:multi_run}).
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains consistently higher in the subglacial system across the time aggregation periods.

Results from  both \alpine{} and \icesheet{} scenarios show that the variability in subglacial outputs decreases with aggregations longer than approximately 1--5 days (Figure~\ref{fig:multi_run}).
Yet, substantial differences in the variability between the model outputs for subglacial and subaerial cases persist in periods of up to $15$ days, longer than which seasonal variations might be conflated.
Note that across longer aggregation times, the timing of peak sediment transport conditions still occur in the natural environment. Yet they are not captured in model outputs(Figure~\ref{fig:model_outs}), despite their potential geomorphic importance.

@@

The different response to water discharge in the model outputs subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}) may in fact cause even greater variations in sediment transport capacity in subglacial channels than presented above.
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}; Table~\ref{tab:Qs}).
Variability increases further due to the application of the exponent occurs before integrating the sediment transport capacity across the channel width, as was done here in evaluating $w\tau$.

>>>>

\begin{keypoints}
\item Rapid water discharge fluctuations in subglacial channels are accommodated by water velocity, not channel size as in subaerial systems.
\item Greater variability in water velocity causes sediment transport capacity to vary more in subglacial channels than in subaerial channels.
\item Significant subglacial sediment transport capacity can persist across a large range of water discharges, impacting  timing of peak events.
\end{keypoints}

\begin{abstract}
  Sediment transport capacity in both subaerial and subglacial channels depends on the shear stress exerted across the channel bottom, which varies with water velocity and channel width.
  In subaerial channels, water discharge variations are accommodated  by flow depth and width changes, along with water velocity.
  However, in subglacial channels  water discharge changes mainly result in velocity variations, as the channel geometry evolves slowly (over days).
  Here, hydrographs from an Alpine glacier and the Greenland Ice Sheet are applied to subglacial and subaerial hydraulics models, revealing greater variability in the inferred sediment transport capacity in subglacial than in subaerial channels.
  Results also show that high subglacial sediment transport rates and water velocities can occur across a wide range of water discharges.
  These findings are discussed in the context of understanding sediment discharge from glaciers with different hydro-climatic forcings and assumptions of channel shape.
\end{abstract}

@@

Periods of millennia or longer, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
Pressurized subglacial water can transport this sediment from underneath glaciers \cite{walder1994,creyts2013,beaud2018,delaney2019}, if it is reachable by the water.

@@

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \cite{leopold1953}.
The change in water depth results in a proportional increase in water velocity, and  the shear stress $\tau$ increases according to Equation~\ref{eq:tau}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels is controlled on the opposing processes of channel opening by frictional heating of water flow on the one hand versus creep closure by ice flow on the other \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days, whereas water discharge can vary more quickly over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, on short periods, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated by changing water velocity $v$ \cite<Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}; >{alley1997}.

@@

Sediment mobilization in subaerial and subglacial channels, therefore, responds differently to changing water discharge.
These differences are implicitly included in a range of models quantifying sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016,pitlick2021}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.

@@

Water discharge data is used from Alpine and ice sheet settings collected downstream of the glacier, assuming no water storage in the proglacial area.
The Alpine site (labeled \alpine{}) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
The water discharge data used here was collected at a $1$\unit{min} interval from May 24, 2014, to October 10, 2014 \cite{felix2022}.

The Leverett Glacier in Greenland (labeled \icesheet{}) serves as the ice sheet setting.
Water discharge was measured roughly $2$\unit{km} downstream from the terminus ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W), at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012 \citeA{tedstone2013}.

@@

\section{Methods}
\label{sect:meth}
The two models described below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both models use the measured discharge to calculate water velocity, shear stress, and width-integrated shear stress, upon which both suspended sediment and bedload transport depend \cite<Figure \ref{fig:cartoon}; >{shields1936}.
Our evaluation in terms of shear stress omits the selection of a sediment transport relationship and a grain-size parameter \cite<e.g.>{shields1936,meyer1948}.

@@

Here, it is assumed that the water is transported through a subglacial channel \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972} beneath a glacier with channel length $l$, with a flat bed and a mean thickness of $h_{ice}$.
The channel  size grows from melt due to frictional heating from water flow and closes due to the creep  of the ice.
The formulation here does not consider the englacial storage of water.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $t$ is time, $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (complete values given in Table \ref{table:vpm}), $g$ is the acceleration due to gravity, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change over $l$, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the mean ice overburden pressure expressed in meter water equivalent ($\rho_w$ is density of water; $\rho_i$ is density of ice), and $n$ is Glen's n \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach equation, the head drop $\Delta h$ is
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{2g} \,f_i\,\frac{v_{g}^{2}}{D_h},
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, $l$ is the channel length, and $v_g=\frac{Q}{S_g}$ is the water flow speed.
% 
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ with
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle, \citeA{hooke1990}). Note that $\beta =\pi$ corresponds to a semi-circular channel and smaller values of $\beta$ result in shallow, wide channels.
This completes the subglacial hydraulic model which is described by the state variables $S_g$ and $\Delta h$.

@@

The models above are applied to proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}).
Outputs of the models represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these scenarios, \alpine{}  is exemplified by Fieschergletscher's relatively thin ice thickness \cite<$h_{ice}$= $225$\,\unit{m}>{grab2021}, low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by Leverett glacier's  thick ice  \cite<$h_{ice}$= $700$\,\unit{m}; >{morlighem2017}, high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).
% However, both scenarios share a discharge with a similar, relative seasonal amplitude.

In the first experiment, the models are  applied to a reference test case for each glacier that assumes  a subglacial channel with $\beta=\frac{\pi}{6}$
(Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities \cite<$\sim\,1.6\,$\unit{m}\,\unit{s}$^{-1}$>{werder2010b,chandler2013} occur for both \alpine{} and \icesheet{}.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with repeatedly applying the maximum observed water discharge of the first $4$ days of the study until there is no change in channel area. 
The subaerial channels of both scenarios have a bed slope of $\nabla z_c = 0.05$ (Table~\ref{table:vpm}).

The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
To quantify the link between sediment transport capacity and water discharge, model outputs are compared to water discharge from the glaciers using Spearman rank correlation, which accounts the ordering of values, but not their magnitude.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.

The second experiment aims to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels across a range of channel shapes and friction factors.
Additionally, we examine the effects of different water discharge variability by smoothing the water discharge over different periods. 
To accomplish this, we examine the variability of each model output by evaluating its standard deviation throughout the season.

The effect of channel shape and roughness is established by running the model with random parameter values of channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_f$; see Table~\ref{table:vpm} for value range).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}\,\unit{s}$^{-1}$ and $2$\,\unit{m}\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}\,\unit{s}$^{-1}$, following measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
Additionally, to be accepted, subglacial model runs can only experience a flotation fraction ($\frac{\Delta h}{2\,h_o}$) of above $1.2$ for less than $2.5$\,\% of the run.

The routine runs until $100$ different parameter combinations for each water discharge smoothing period are accepted with the conditions described above.
Here, the water discharge input at a given time is the average over the smoothing period before that time, ranging from $15$\,\unit{min} up to $15$\,\unit{days}.

\section{Results}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.png}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows show examples where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the  peak melt event denoted by the shaded area in panels e--h, with an arbitrary y-axis.
    % Note that the x and y axes are different for \alpine{} and \icesheet{}.
  }
  \label{fig:model_outs}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig3.png}
  \caption{
    Relationship between water discharge and normalized velocity, shear stress, and width-integrated shear stress for \alpine{} (a-c) and \icesheet{} (d-f).
    Variables on the x-axis have been normalized to mean values.
    $R_g$ ($R_f$) shows the Spearman rank correlations for the subglacial (subaerial) outputs.
    Plots are shown with discharge and outputs at $15$\unit{min} intervals.
  }
  \label{fig:Qw_vari}
\end{figure}

Examining the relationship between water discharge and the model outputs, different relationships emerge for the subaerial and subglacial cases due to the hysteresis in channel size in the subglacial model (Equation~\ref{eq:dS_dt}).
Because there is no history dependence in  subaerial channels, each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari}).

Conversely, the history dependence on channel size in subglacial channels means different outputs occur across a range of discharges.
For instance, in subglacial channels in \alpine{}, very high seasonal water velocity values and shear stresses can occur from a low water discharge  ($\sim\,4$\,\unit{m}$^3$\,\unit{s}$^{-1}$) to the maximum water discharge at over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ (Figure~\ref{fig:Qw_vari}\,a).
In \icesheet{}, water velocities close to the seasonal mean value can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
When considering the subglacial channel's evolving width, width-integrated shear stress  across the channel generally increases with water discharge, with improved rank correlation compared to water velocity or shear stress (\alpine{}, Figure~\ref{fig:Qw_vari} a--c).
Yet even the width-integrated shear stress  can vary substantially, with the highest values occurring at water discharge values ranging from roughly $11$ \,\unit{m}\,\unit{s}$^{-1}$ to over $17$ \,\unit{m}\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet{} scenario, where the hydrograph has less diurnal variability (Figure~\ref{fig:Qw_vari} c, f).

The different response of subaerial and subglacial model outputs to water discharge impacts the timing of peaks in sediment transport capacity, and thus its  relationship with water discharge.
In the subglacial channel, peaks in $w_g\tau_g$ generally occur when water discharge increases at the fastest rate, but before the maximum water discharge (Figure~\ref{fig:model_outs}\, d and h).
By contrast, in the subaerial channel,   $w_f\tau_f$  peaks when water discharge is highest.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig4.pdf}
  \caption{Standard deviation of model outputs for different smoothing times with different subglacial and subaerial channel shapes and friction factors.
    Shaded areas denote the range of standard deviations from the accepted $100$ parameter combinations (Section~\ref{sect:imp}).
    Solid lines denote  mean value of standard deviations.
    Markers show smoothing periods ($15$\,\unit{min}, $1$\,\unit{hr}, $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d}).
  }
  \label{fig:multi_run}
\end{figure}

The second numerical experiment aims to evaluate the response of variability in model outputs to a range of channel shapes, friction factors, and water discharge variability.
Across the range of parameters  examined (Section\,\ref{sect:imp}), variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one in both \alpine{} and \icesheet{}.
This suggests that there is greater intermittency and variability in sediment transport in subglacial systems compared to subaerial ones, especially in supply-limited systems where sediment could be exhausted or mobilized over diurnal discharge cycles.
Considering the width integrated shear stress, in \alpine{} the variability is comparable between the subaerial and subglacial systems (shown by overlap of plots), due to the highly variable water discharge, largely due to diurnal variations (Figure~\ref{fig:multi_run}).
In contrast, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains consistently higher in the subglacial system across the time smoothing periods.

Results from  both \alpine{} and \icesheet{} scenarios show that the variability in subglacial outputs decreases with smoothing time longer than approximately 1--5 days (Figure~\ref{fig:multi_run}).
Yet, substantial differences in the variability between the model outputs for subglacial and subaerial cases persist in periods of up to $15$ days, longer than which seasonal variations might be conflated.
Note that across longer smoothing times, the timing of peak sediment transport conditions still occur in the natural environment. Yet they are not captured in model outputs(Figure~\ref{fig:model_outs}), despite their potential geomorphic importance.

@@

Increased variability in subglacial sediment discharge capacity makes it more difficult to differentiate the role of long term climatic controls on sediment transport with shorter term hydrological variations.
The different response to water discharge variations in the model outputs subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}) may in fact cause even greater variations in sediment transport capacity in subglacial channels than presented above.
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}; Table~\ref{tab:Qs}).

For sufficiently short periods, channel width and wetted area are fixed in the subglacial system (akin to pipe flow) so that the response of width-integrated shear stress is $w_g\, \tau_{g}\, \propto\,  Q^{2}$ \cite<Table\,\ref{tab:eqs1}; >{alley1997}.
\ref{sect:scaling} provides a complete description of sediment transport capacity in different drainage conditions.
Note that in pipe flow conditions, maximum sediment transport would coincide with peak water discharge, unlike here where evolving channel size here alters the timing (Figure \ref{fig:model_outs} and \ref{fig:Qw_vari}).
R\"othlisberger channels in steady state with water discharge, the width-integrated shear stress response to water discharge is $ w_g\,\tau_{g}\, \propto\,  Q^{\sim\frac{4}{5}}$ (Table\,\ref{tab:eqs1}).
This relationship suggests that when discharge varies more slowly than channel shape subglacial channels in fact have less variable sediment transport capacity than subaerial ones.
Yet, subglacial channels rarely operate in steady state \cite{gimbert2016}.
